The ability of single genes vs full genomes to resolve time and space in outbreak analysis

### 2019-10-15 genomic-horizon.tex
Final final update to the manuscript.

The combination of decreasing cost of sequencing and the unparalleled insight it offers have led to the adoption of pathogen genetic sequencing as one of the most effective tools in a modern epidemiologist's toolkit.
When coupled with sophisticated models of evolution, pathogen sequences can be used to look into epidemiological features such as cryptic transmission \cite{faria_establishment_2017}, migration \cite{lemey_bayesian_2009,lemey_unifying_2014}, and origins \cite{smith_origins_2009} of infectious diseases, amongst others.
Pathogen sequences also contain information about past temporal dynamics before sequence data have been collected \cite{raghwani_origin_2012} due to the pattern of shared and unique mutations inherited from preceding generations.
Molecular phylogenetic approaches rely on decoding these patterns of shared mutations into a nested graph known as the phylogenetic tree.
Pathogens often have short generation times and some, like RNA viruses, also possess polymerases with low replication fidelity such that mutations are generated at a rapid pace \cite{drummond_measurably_2003,biek_measurably_2015} leading to fast differentiation of pathogen lineages at the genetic level as they spread.
With appropriate sampling and information (``metadata'') about sequences, historic population dynamics can be inferred and quantified from pathogen phylogenies.
Changes in pathogen population sizes over time \cite{pybus_integrated_2000}, inference of unobserved ancestral states \cite{lemey_bayesian_2009,dudas_mers-cov_2018}, correlates of processes \cite{faria_simultaneously_2013,lemey_unifying_2014,dudas_virus_2017}, and overall phylodynamic \cite{grenfell_unifying_2004} patterns can be inferred from molecular phylogenies and used to understand patterns of pathogen transmission at a number of scales.

Before widespread adoption of high-throughput sequencing limitations and costs led to amplification and sequencing of short fragments of pathogen genomes \cite{jin_genetic_1999,jin_proposal_2005}.
These subgenomic fragments were often chosen for their diversity, such as viral surface glycoproteins that experience selective pressures from vertebrate immune systems, or their utility, such as routine sequencing of HIV \textit{pol} gene to test for drug resistance \cite{kaye_phylogenetic_2008,rhee_human_2003}.
Whilst subgenomic fragments of pathogens are very accurate and specific as diagnostic markers and informative about long-term evolution, their length (dictated by the compromise between information content and ease of sequencing) limits their utility in detailed molecular epidemiology investigations, for example during outbreaks \cite{wohl_co-circulating_2018}, as only mutations occurring within the small region of the genome that is sequenced are available for phylogenetic inference.

Molecular clocks have been particularly useful in molecular epidemiology, where the accumulation of mutations between sequences is used as a noisy approximation for elapsed time, given either times of events in the phylogeny (sequence dates or dates of common ancestors) or a previously determined molecular clock rate.
Generally, neutral pathogen variation at the nucleotide level ebbs and flows under the forces of population genetics, unlike beneficial or deleterious variation which tends to either fix or be purged rapidly, respectively.
Due to their random and discrete nature, mutations are modelled as a Poisson process \cite{yang_computational_2006} where the waiting time $t$ for observing a mutation at a single site is exponentially distributed with evolutionary rate parameter $R$.
The probability of observing 0 mutations at a single site after time $t$ is $e^{-Rt}$ and the probability of at least one mutation is therefore $1-e^{-Rt}$.
Higher evolutionary rates \textit{R} or waiting times \textit{t} result in higher probabilities of observing at least one mutation at the site in question.
Since sites are assumed to evolve independently, the probability of observing at least one mutation across \textit{L} sites is
\begin{equation}
  P = 1-e^{-RLt},
\end{equation}
where \textit{RL} is expressed in substitutions per year (rate in substitutions per site per year multiplied by number of sites).
Since the probability of observing at least one mutation changes depending on waiting time (and \textit{vice versa}) we instead focus on the mean waiting time until at least one mutation appears.
The mean of an exponential distribution is $\lambda^{-1}$ where $\lambda$ in our case is \textit{RL}, such that mean waiting time $\bar{t}$ under a given evolutionary rate \textit{R}, and sequence length \textit{L} becomes
\begin{equation}
  \bar{t} = \frac{1}{RL}
  \label{horizon}
\end{equation}
When the evolutionary rate \textit{R} or sequence length \textit{L} are low mean waiting times $\bar{t}$ are lengthened and \textit{vice versa}.
It also suggests a worrying relationship between \textit{R}, \textit{L}, and $\bar{t}$ - a reduction in either \textit{R} or \textit{L} leads to a reciprocal reduction in $\bar{t}$, such that reducing the number of sites by 10\%, for example, requires a $\frac{1}{0.9} = 1.11(1)$-fold increase in evolutionary rate to maintain the same temporal resolution $\bar{t}$ or risk increasing the waiting time for a mutation by the same amount (\textit{i.e.} $\approx$11\%, or 0.9 of the original resolution available).
This gets worse reciprocally, such that halving the sites requires doubling the evolutionary rate, using 10\% of the sites requires a ten-fold increase in evolutionary rate, \textit{etc}.

Since both maximum plausible evolutionary rates $R$ and genome length $G$ are largely dictated by deleterious mutation load \cite{gago_extremely_2009}, neither quantity will vary substantially for a given pathogen, though individually $R$ and $G$ can vary substantially, where for example viruses have high $R$ and low $G$ on average, and bacteria have higher $G$ but lower $R$.
Sequencing recovers some fraction $f$ of the genome length $G$ ($L=Gf$) for analyses, and sequencing complete genomes ($f=1.0; L = G$) is the best possible scenario, since sequencing any shorter region requires the evolutionary rate to increase by a factor of $\frac{1}{f}$, which even if $L = 0.4 G$ means the evolutionary rate would have to be 2.5 times faster in the remaining 0.4 of the genome to be able to record information in the form of mutations at the same speed as complete genomes.
The message of our manuscript, at least as far as densely sampled infectious disease outbreaks go, is that the task of sequencing a complete pathogen genome will rarely be as miserable a task as analysing a fraction of one.

In this study we show this by quantifying how much information relevant to phylodynamic analysis is lost when shorter genomic regions are used instead of full genomes.
By focusing our attention on a subset (600 sequences) of a well-characterised genomic sequence data (comprised of $>$1600 viral genomes) set derived from the West African Ebola virus epidemic of 2013-2015 \cite{dudas_virus_2017} we estimate loss in precision and accuracy of molecular clock models and phylogeographic inference methods when only the glycoprotein gene (GP), a region representing just 10\% of the viral genome, is analysed despite GP evolving at rates faster than the genomic average.
Our methods rely on masking tip dates and locations for 60 (10\%) of the sequences in a classic training-testing split, where we re-infer these parameters as latent variables using Markov chain Monte Carlo (MCMC).
We show that this reduction in data not only leads to severe mixing issues in MCMC analyses by removing the constraints additional data impose on plausible parameter space without adding restrictive priors to compensate, but can also result in unreliable tip date and location inference.
Despite achieving much better temporal resolution when using complete viral genomes we still find residual error caused by inherent randomness of mutations which is close to theoretical expectations (Eq \ref{horizon}).
We refer to this as the temporal horizon, \textit{i.e.} a temporal resolution limit where population processes occurring at a rate faster than the rate at which mutations enter and are observed in a population will not be captured with high fidelity, even with genome sequences.

@@

Figure \ref{trees} shows the reconstructed phylogenies in substitution space (right) and time space (left) for 600 complete Ebola virus genomes (top) or just GP sequences (bottom).
Although higher levels of divergence are observed in the GP dataset, as seen from tree height, the differences in the number of non-polytomic nodes between genomic and GP data are clear, indicating substantially better resolution in disentangling the exact relationships between lineages in the former.
Figure \ref{embedding} shows where in the better resolved maximum likelihood phylogeny of genome sequences the mutations that occurred in just the region spanning the GP gene are located, with continuous blocks of colours corresponding to regions of the tree that would become collapsed if only the GP data were used and the tree were inferred correctly.
Internal branches of a phylogeny correspond to hypotheses of common ancestry and in the case of GP only 42 internal nodes are identified in the maximum likelihood phylogeny compared to 210 internal nodes for complete genomes.
The one aspect of the West African epidemic that can be inferred from both GP and genome phylogenies is that the virus' origins lie in Guinea, but details of its onwards spread are largely lost in the GP phylogeny.
Genomic data, on the other hand, despite a reduction from over 1600 sequences described in the original study \cite{dudas_virus_2017} down to just 600, still contain information about the role of Sierra Leone's epidemic in maintaining transmission across the region through both endemic proliferation of lineages and their spread to neighbouring countries.

Unlike maximum likelihood phylogenies where branch lengths are directly proportional to the expected number of substitutions, branch lengths in temporal phylogenies are usually smoothed out by the fact that a range of dates are compatible with a given number of mutations on a branch.
Thus even large polytomies can be resolved into a branching structure derived from the tree prior, albeit without much support for any given configuration.
%%%%
So despite the maximum likelihood of GP exhibiting a lot of polytomies (Figure \ref{trees}) the corresponding temporal phylogeny (left) does not, though it is more star-shaped than its genome equivalent, with long external branches as well as numerous nodes with less than 0.10 posterior support (marked by grey crosses).
%%%%
Though the trees were too large to estimate robust tree distance statistics such as subtree prune and regraft (SPR) distances, Robinson-Foulds distances are not, and are smaller between the GP and genome maximum likelihood trees (188) than between Bayesian timetrees (1068).
There are also noticeable differences in total tree length, and whereas it is entirely expected that the maximum likelihood tree of GP should be larger (0.08076 substitutions/site) than the genome substitution phylogeny (0.06782 subs/site) due to the former's faster rate of evolution, the tree length of time trees differ 2-fold - 80.299 years for genome versus 173.018 years for GP.
There is also a noticeable degree of branch clustering by country in the GP temporal phylogeny, possibly caused by proximity of locations within country, which in the absence of genetic information cannot be resolved to the same degree as with genomic data.

In contrast to the maximum likelihood phylogeny of GP on the right (Figure \ref{trees}) its corresponding temporal phylogeny on the left exhibits a reconstruction of the West African epidemic largely consistent with what has been established previously \cite{dudas_virus_2017}.
This is likely to be caused by the combined effects of two sources of information.
First, additional information is added by specifying the collection dates for sequences, which might exclude certain topologies from being considered during MCMC on account of the relatively small effective population size of Ebola virus in West Africa.
Second, the generalised linear model approach to inferring migration is information-rich as it provides over 3000 possible parameter values (pairwise migration rates between locations) per predictor matrix, and thus if a few branches are strongly selecting for a ``correct'' predictor matrix to be included in the migration model that predictor matrix can then be used to determine the likely locations of branches for which less information is available.
However, a simpler continuous time Markov chain model where each individual pairwise migration rate is inferred individually in a maximum likelihood framework exhibits broadly similar patterns too (Figure \ref{TTtrees}).

@@

Inferring masked tip dates from 10\% of the sequences (Figure \ref{dates}) is an intuitive way to show both the inherent noisiness of molecular clock estimates, as reflected in the width of 95\% highest posterior density intervals for inferred dates, and the differences in temporal resolution between GP and genome alignments.
True collection dates for genomes are mostly (56 out of 60, corresponding to a coverage probability of 0.93) within the 95\% highest posterior density of estimated dates, and the mean absolute error is $\approx$22 days across all masked tips.
In contrast, the 95\% HPDs for inferred dates in the GP dataset capture more of the true dates (58 out of 60, coverage probability $\approx$0.96) at the cost of markedly reduced precision, with mean absolute error going up to $\approx$106 days or $\sim$3.5 months.
Despite having lower coverage probability, more precise date estimates are derived from complete genomes with an average 95\% HPD width of $\approx$102 days, compared to $\approx$458 days for GP.
Another way of thinking about where the loss of information occurs is to consider root-to-tip against tip date regressions shown in Figure \ref{rtt}, where waiting times for mutations are too long to estimate the slope of the regression reliably, as every new mutation is seen across sequences collected over a longer interval of time.
Observed errors (Figure \ref{dates}, but also Figure \ref{TTdates} for maximum likelihood equivalent) for both datasets are very close to theoretical expectations calculated using Equation \ref{horizon}: 22 (observed) versus 20 (expected) days for Ebola virus genomes, and 106 (observed) versus 113 (expected) days for GP.
Also note that for many tips in the GP data set independent Markov chains in some cases converged on, and in other cases sampled from, different distributions for masked tip dates (\textit{i.e.} local maxima) resulting in multi-peaked posterior samples after combining independent analyses.

@@

Differences between genomic and GP datasets are clear and dramatic when looking at both phylogenies (Figure \ref{trees}) and masked date inference (Figure \ref{dates}), but less pronounced when trying to infer the location of a masked tip (Figure \ref{locations}).
Although locations are correctly inferred more often and with greater support in genomic sequences compared to just the GP gene, there are numerous tips whose locations are not correctly inferred even from genome sequences (Figure \ref{locations} and Figure \ref{TTlocations} for maximum likelihood equivalent).
This might reflect the nature of these parameters of interest, since phylogenies and date inference ultimately draw information from mutation accumulation via relatively straightforward models of sequence evolution with limited parameter space.
In contrast, migration processes are far more complicated and nuanced without a \textit{de facto} standard for modelling, though continuous time Markov chain (CTMC) approaches are widely used \cite{lemey_bayesian_2009} with most advanced methods relying on generalised linear models without excessive over-parameterisation \cite{faria_simultaneously_2013,lemey_unifying_2014,dudas_virus_2017}.
Despite the lack of strong contrast in power to infer masked locations between genomes and GP sequences, cross entropies indicate better performance with complete genomes (6054.631 nats) than with GP (9905.726 nats).

Similarly, locations are inferred correctly more often with complete genomes than with GP sequences, where the maximum probability location (\textit{i.e.} the model's best guess) matches the truth, where using complete genomes results in 0.540 probability of guessing correctly compared to 0.286 probability for GP (for a calibration of both models see Figure \ref{calibration}).
The model makes these guesses with more certainty too, where the mean probability of the true location is 0.482 with genomes and 0.219 with GP, and mean probability of best guess (\textit{i.e.} maximum probability) is 0.680 and 0.396, respectively.
We also calculated the great circle distances between the population centroids of true and each predicted location weighed by probability, which should ideally be 0.0 (0 km distance multiplied by probability of 1.0).
The mean of these distances across masked tips are 75.886 kilometres for genomes compared to 164.309 km for GP sequences.

In addition to assessing how well tip locations can be inferred from genetic information we also looked at how well historical patterns were reconstructed from sequence data.
To accomplish this we looked at the posterior distribution of ancestral locations of lineages that gave rise to four sequences in the data.
The four lineages were chosen for their well-characterised histories in the broader epidemic as well as complexity of migration.
Note that when we describe the movements of these lineages they are referred to by the strains that would eventually descend from them, but it should in no way be interpreted as a single virus or patient moving around the region throughout the epidemic.
Instead, all four strains share a number of common ancestors (the tree that would describe their relationships can be represented as (((`14859\_EMLK', `MK3462'),`EM\_004422'),`PL5294');) whose locations can be inferred from the geographic distribution of their relatives and descendants.

Of the four representative sequenced viruses chosen, three (`14859\_EMLK', `MK3462', `EM\_004422') are descended from the viral lineage that jumped across the border from Gu\'{e}ck\'{e}dou prefecture in Guinea into Kailahun district in Sierra Leone.
The common ancestor of two of those (`14859\_EMLK' and `MK3462') continued onto Kenema district in Sierra Leone from Kailahun district (also Sierra Leone).
While the lineage/transmission chain that eventually gave rise to strain `14859\_EMLK' continued onto Conakry prefecture in Guinea and back-spilled into Sierra Leone's Kambia district (where the descendent strain `14859\_EMLK' was collected \cite{arias_rapid_2016}) right across the border later in the epidemic, the lineage/transmission chain that was ancestral to `MK3462' stayed in Sierra Leone for the remainder of the epidemic, found itself moved westwards towards Freetown (Western Urban and Western Rural districts) and finally jumping to Bombali district, where its descendent strain was collected from a patient \cite{arias_rapid_2016}.
Unlike all the other three lineages, the transmission chain/lineage (``lineage A'' \cite{carroll_temporal_2015}) that would leave strain `PL5294' as a descendent is thought to have been largely restricted to the environs of Conakry prefecture in Guinea.
An older ancestor of strain `PL5294' had migrated from Gu\'{e}ck\'{e}dou prefecture into Conakry prefecture on the other side of the country relatively early in the epidemic, where its descendents circulated for a large portion of the West African epidemic, though unlike a lot of its relatives the direct ancestor that would eventually give rise to strain `PL5294' spilled over into Kambia district of Sierra Leone \cite{arias_rapid_2016}.
Finally, the transmission chain/lineage that would eventually give rise to strain `EM\_004422' had a tumultous history in the region.
Strains `EM\_004422', `14859\_EMLK' and `MK3462' all shared the same common ancestor in Kailahun district of Sierra Leone, but the progenitor of strain `EM\_004422', unlike its relatives, made a jump to Liberia (Lofa and Montserrado counties), from where \textit{its} descendants spilled back into Macenta prefecture in Guinea.
This transmission chain would later leave descendants (of which `EM\_004422' is representative) that jumped to neighbouring Kissidougou prefecture (also Guinea) and where strain `EM\_004422' was collected \cite{carroll_temporal_2015}.

The histories of the lineages that gave rise to these four tips are, for the most part, reconstructed from both GP sequences and genomes consistently (Figure \ref{trace}), likely as a result of additional information brought in by specifying tip dates and their collection locations.
Genomic data tend to concentrate the probability mass towards a single location at any given time, in contrast to GP sequences where several locations can be considered with non-negligible probabilities at numerous time points (Figure \ref{trace} and Figure \ref{trace_entropy}) and where timing of ancestral migration events is considerably more diffuse or even substantially off (\textit{i.e.} Figure \ref{trace}A and D).
What is even more apparent is that without the additional information available when using complete genomes and without aiding the sampling with strongly informative priors MCMC explores a wider variety of low-probability migration paths, as indicated by maps on the right of each plot in Figure \ref{trace}.
In the case of the ancestral lineage of `EM\_004422', for example, a series of migrations through distant Conakry (western Guinea) are reconstructed with relatively high confidence from GP sequences, compared to shorter distance migrations that run through neighbouring Liberia reconstructed from genomes.

>>>>

The combination of decreasing cost of sequencing and the unparalleled insight it offers have led to the adoption of pathogen genetic sequencing as one of the most effective tools in a modern epidemiologist's toolkit.
When coupled with sophisticated models of evolution pathogen sequences can be used to look into epidemiological features such as cryptic transmission \cite{faria_establishment_2017}, migration \cite{lemey_bayesian_2009,lemey_unifying_2014}, and origins \cite{smith_origins_2009} of infectious diseases amongst others.
Pathogen sequences also contain information about past temporal dynamics before sequence data have been collected \cite{raghwani_origin_2012} due to the pattern of shared and unique mutations inherited from preceding generations.
Molecular phylogenetic approaches rely on decoding these patterns of shared mutations into a nested graph known as the phylogenetic tree.
Pathogens often have short generation times and some, like RNA viruses, also possess polymerases with low replication fidelity such that mutations are generated at a rapid pace \cite{drummond_measurably_2003,biek_measurably_2015} leading to fast differentiation of pathogen lineages at the genetic level as they spread.
With appropriate sampling and information (``metadata'') about sequences historic population dynamics can be inferred and quantified from pathogen phylogenies.
Changes in pathogen population sizes over time \cite{pybus_integrated_2000}, inference of unobserved ancestral states \cite{dudas_mers-cov_2018}, correlates of processes \cite{faria_simultaneously_2013,dudas_virus_2017}, and overall phylodynamic \cite{grenfell_unifying_2004} patterns can be inferred from molecular phylogenies and used to understand patterns of pathogen transmission at a number of scales.

Before widespread adoption of high-throughput sequencing limitations and costs led to amplification and sequencing of short fragments of pathogen genomes \cite{jin_genetic_1999,jin_proposal_2005}.
These subgenomic fragments were often chosen for their diversity such as viral surface glycoproteins that experience selective pressures from vertebrate immune systems or their utility such as routine sequencing of HIV \textit{pol} gene to test for drug resistance \cite{kaye_phylogenetic_2008,rhee_human_2003}.
Whilst subgenomic fragments of pathogens are very accurate and specific as diagnostic markers and informative about long-term evolution their length (dictated by the compromise between information content and ease of sequencing) limits their utility in detailed molecular epidemiology investigations for example during outbreaks \cite{wohl_co-circulating_2018} as only mutations occurring within the small region of the genome that is sequenced are available for phylogenetic inference.

Molecular clocks have been particularly useful in molecular epidemiology where the accumulation of mutations between sequences is used as a noisy approximation for elapsed time, given either times of events in the phylogeny (sequence dates or dates of common ancestors) or a previously determined molecular clock rate.
Generally neutral pathogen variation at the nucleotide level ebbs and flows under the forces of population genetics unlike beneficial or deleterious variation which tends to either fix or be purged rapidly, respectively.
Due to their random and discrete nature mutations are modelled as a Poisson process \cite{yang_computational_2006} where the waiting time $t$ for observing a mutation at a single site is exponentially distributed with evolutionary rate parameter $R$.
The probability of observing 0 mutations at a single site after time $t$ is $e^{-Rt}$ and the probability of at least one mutation is therefore $1-e^{-Rt}$.
Higher evolutionary rates \textit{R} or waiting times \textit{t} result in higher probabilities of observing at least one mutation at the site in question.
Since sites are assumed to evolve independently the probability of observing at least one mutation across \textit{L} sites is
\begin{equation}
  P = 1-e^{-RLt},
\end{equation}
where \textit{RL} is expressed in substitutions per year (rate in substitutions per site per year multiplied by number of sites).
Since the probability of observing at least one mutation changes depending on waiting time (and \textit{vice versa}) we instead focus on the mean waiting time until at least one mutation appears.
The mean of an exponential distribution is $\lambda^{-1}$ where $\lambda$ in our case is \textit{RL} such that mean waiting time $\bar{t}$ under a given evolutionary rate \textit{R} and sequence length \textit{L} becomes
\begin{equation}
  \bar{t} = \frac{1}{RL}
  \label{horizon}
\end{equation}
When the evolutionary rate \textit{R} or sequence length \textit{L} are low mean waiting times $\bar{t}$ are lengthened and \textit{vice versa}.
It also suggests a worrying relationship between \textit{R}, \textit{L}, and $\bar{t}$ - a reduction in either \textit{R} or \textit{L} leads to a reciprocal reduction in $\bar{t}$ such that reducing the number of sites by 10\%, for example, requires a $\frac{1}{0.9} = 1.11(1)$-fold increase in evolutionary rate to maintain the same temporal resolution $\bar{t}$ or risk increasing the waiting time for a mutation by the same amount (\textit{i.e.} $\approx$11\% or 0.9 of the original resolution available).
This gets worse reciprocally such that halving the sites requires doubling the evolutionary rate, using 10\% of the sites requires a ten-fold increase in evolutionary rate, \textit{etc}.

Since both maximum plausible evolutionary rates $R$ and genome length $G$ are largely dictated by deleterious mutation load neither quantity will vary substantially for a given pathogen though individually $R$ and $G$ can vary substantially where for example viruses have high $R$ and low $G$ on average and bacteria have higher $G$ but lower $R$.
Sequencing recovers some fraction $f$ of the genome length $G$ ($L=Gf$) for analyses and sequencing complete genomes ($f=1.0; L = G$) is the best possible scenario since sequencing any shorter region requires the evolutionary rate to increase by a factor of $\frac{1}{f}$ which even if $L = 0.4 G$ means the evolutionary rate would have to be 2.5 times faster in the remaining 0.4 of the genome to be able to record information in the form of mutations at the same speed as complete genomes.
The message of our manuscript, at least as far as densely sampled infectious disease outbreaks go, is that the task of sequencing a complete pathogen genome will rarely be as miserable a task as analysing a fraction of one.

In this study we show this by quantifying how much information relevant to phylodynamic analysis is lost when shorter genomic regions are used instead of full genomes.
By focusing our attention on a subset (600 sequences) of a well-characterised genomic sequence data (comprised of $>$1600 viral genomes) set derived from the West African Ebola virus epidemic of 2013-2015 \cite{dudas_virus_2017} we estimate loss in precision and accuracy of molecular clock models and phylogeographic inference methods when only the glycoprotein gene (GP), a region representing just 10\% of the viral genome, is analysed despite GP evolving at rates faster than the genomic average.
Our methods rely on masking tip dates and locations for 60 (10\%) of the sequences in a classic training-testing split where we re-infer these parameters as latent variables using Markov chain Monte Carlo (MCMC).
We show that this reduction in data not only leads to severe mixing issues in MCMC analyses by removing the constraints additional data impose on plausible parameter space without adding restrictive priors to compensate, but can also result in unreliable tip date and location inference.
Despite achieving much better temporal resolution when using complete viral genomes we still find residual error caused by inherent randomness of mutations which is close to theoretical expectations (Eq \ref{horizon}).
We refer to this as the temporal horizon, \textit{i.e.} a temporal resolution limit where population processes occurring at a rate faster than the rate at which mutations enter and are observed in a population will not be captured with high fidelity even with genome sequences.

@@

Figure \ref{trees} shows the reconstructed phylogenies in substitution space (right) and time space (left) for 600 complete Ebola virus genomes (top) or just GP sequences (bottom).
Although higher levels of divergence are observed in the GP dataset, as seen from tree height, the differences in the number of non-polytomic nodes between genomic and GP data are clear, indicating substantially better resolution in disentangling the exact relationships between lineages in the former.
Figure \ref{embedding} shows where in the better resolved maximum likelihood phylogeny of genome sequences the mutations that occurred in just the region spanning the GP gene are located with continuous blocks of colours corresponding to regions of the tree that would become collapsed if only the GP data were used and the tree were inferred correctly.
Internal branches of a phylogeny correspond to hypotheses of common ancestry and in the case of GP only 42 internal nodes are identified in the maximum likelihood phylogeny compared to 210 internal nodes for complete genomes.
The one aspect of the West African epidemic that can be inferred from both GP and genome phylogenies is that the virus' origins lie in Guinea but details of its onwards spread are largely lost in the GP phylogeny.
Genomic data, on the other hand, despite a reduction from over 1600 sequences described in the original study \cite{dudas_virus_2017} down to just 600 still contain information about the role of Sierra Leone's epidemic in maintaining transmission across the region through both endemic proliferation of lineages and their spread to neighbouring countries.

Unlike maximum likelihood phylogenies where branch lengths are directly proportional to the expected number of substitutions branch lengths in temporal phylogenies are usually smoothed out by the fact that a range of dates are compatible with a given number of mutations on a branch.
Thus even large polytomies can be resolved into a branching structure derived from the tree prior albeit without much support for any given configuration.
%%%%
So despite the maximum likelihood of GP exhibiting a lot of polytomies (Figure \ref{trees}) the corresponding temporal phylogeny (left) does not, though it is more star-shaped than its genome equivalent with long external branches as well as numerous nodes with less than 0.10 posterior support (marked by grey crosses).
%%%%
Though the trees were too large to estimate robust tree distance statistics such as subtree prune and regraft (SPR) distances, Robinson-Foulds distances are not, and are smaller between the GP and genome maximum likelihood trees (188) than between Bayesian timetrees (1068).
There are also noticeable differences in total tree length and whereas it is entirely expected that the maximum likelihood tree of GP should be larger (0.08076 substitutions/site) than the genome substitution phylogeny (0.06782 subs/site) due to the former's faster rate of evolution the tree length of time trees differ 2-fold - 80.299 years for genome versus 173.018 years for GP.
There is also a noticeable degree of branch clustering by country in the GP temporal phylogeny possibly caused by proximity of locations within country which in the absence of genetic information cannot be resolved to the same degree as with genomic data.

In contrast to the maximum likelihood phylogeny of GP on the right (Figure \ref{trees}) its corresponding temporal phylogeny on the left exhibits a reconstruction of the West African epidemic largely consistent with what has been established previously \cite{dudas_virus_2017}.
This is likely to be caused by the combined effects of two sources of information.
First, additional information is added by specifying the collection dates for sequences which might exclude certain topologies from being considered during MCMC on account of the relatively small effective population size of Ebola virus in West Africa.
Second, the generalised linear model approach to inferring migration is information-rich as it provides over 3000 possible parameter values (pairwise migration rates between locations) per predictor matrix and thus if a few branches are strongly selecting for a ``correct'' predictor matrix to be included in the migration model that predictor matrix can then be used to determine the likely locations of branches for which less information is available.
However, a simpler continuous time Markov chain model where each individual pairwise migration rate is inferred individually in a maximum likelihood framework exhibits broadly similar patterns too (Figure \ref{TTtrees}).

@@

Inferring masked tip dates from 10\% of the sequences (Figure \ref{dates}) is an intuitive way to show both the inherent noisiness of molecular clock estimates as reflected in the width of 95\% highest posterior density intervals for inferred dates and the differences in temporal resolution between GP and genome alignments.
True collection dates for genomes are mostly (56 out of 60, corresponding to a coverage probability of 0.93) within the 95\% highest posterior density of estimated dates and the mean absolute error is $\approx$22 days across all masked tips.
In contrast the 95\% HPDs for inferred dates in the GP dataset capture more of the true dates (58 out of 60, coverage probability $\approx$0.96) at the cost of markedly reduced precision with mean absolute error going up to $\approx$106 days or $\sim$3.5 months.
Despite having lower coverage probability more precise date estimates are derived from complete genomes with an average 95\% HPD width of $\approx$102 days compared to $\approx$458 days for GP.
Another way of thinking about where the loss of information occurs is to consider root-to-tip against tip date regressions shown in Figure \ref{rtt} where waiting times for mutations are too long to estimate the slope of the regression reliably as every new mutation is seen across sequences collected over a longer interval of time.
Observed errors (Figure \ref{dates} but also Figure \ref{TTdates} for maximum likelihood equivalent) for both datasets are very close to theoretical expectations calculated using Equation \ref{horizon}: 22 (observed) versus 20 (expected) days for Ebola virus genomes and 106 (observed) versus 113 (expected) days for GP.
Also note that for many tips in the GP data set independent Markov chains in some cases converged on, and in other cases sampled from, different distributions for masked tip dates (\textit{i.e.} local maxima) resulting in multi-peaked posterior samples after combining independent analyses.

@@

Differences between genomic and GP datasets are clear and dramatic when looking at both phylogenies (Figure \ref{trees}) and masked date inference (Figure \ref{dates}) but less pronounced when trying to infer the location of a masked tip (Figure \ref{locations}).
Although locations are correctly inferred more often and with greater support in genomic sequences compared to just the GP gene there are numerous tips whose locations are not correctly inferred even from genome sequences (Figure \ref{locations} and Figure \ref{TTlocations} for maximum likelihood equivalent).
This might reflect the nature of these parameters of interest since phylogenies and date inference ultimately draw information from mutation accumulation via relatively straightforward models of sequence evolution with limited parameter space.
In contrast, migration processes are far more complicated and nuanced without a \textit{de facto} standard for modelling though continuous time Markov chain (CTMC) approaches are widely used with most advanced methods relying on generalised linear models without excessive over-parameterisation.
Despite the lack of strong contrast in power to infer masked locations between genomes and GP sequences cross entropies indicate better performance with complete genomes (6054.631 nats) than with GP (9905.726 nats).

Similarly, locations are inferred correctly more often with complete genomes than with GP sequences where the maximum probability location (\textit{i.e.} the model's best guess) matches the truth.
Specifically, using complete genomes results in 0.540 probability of guessing correctly compared to 0.286 probability for GP (for a calibration of both models see Figure \ref{calibration}).
The model makes these guesses with more certainty too where the mean probability of the true location is 0.482 with genomes and 0.219 with GP and mean probability of best guess (\textit{i.e.} maximum probability) is 0.680 and 0.396, respectively.
We also calculated the great circle distances between the population centroids of true and each predicted location weighed by probability which should ideally be 0.0 (0 km distance multiplied by probability of 1.0).
The mean of these distances across masked tips are 75.886 kilometres for genomes compared to 164.309 km for GP sequences.

In addition to assessing how well tip locations can be inferred from genetic information we also looked at how well historical patterns were reconstructed from sequence data.
To accomplish this we looked at the posterior distribution of ancestral locations of lineages that gave rise to four sequences in the data.
The four lineages were chosen for their well-characterised histories in the broader epidemic as well as complexity of migration.
Note that when we describe the movements of these lineages they are referred to by the strains that would eventually descend from them but it should in no way be interpreted as a single virus or patient moving around the region throughout the epidemic.
Instead, all four strains share a number of common ancestors (the tree that would describe their relationships can be represented as (((`14859\_EMLK', `MK3462'),`EM\_004422'),`PL5294');) whose locations can be inferred from the geographic distribution of their relatives and descendants.

Of the four representative sequenced viruses chosen three (`14859\_EMLK', `MK3462', `EM\_004422') are descended from the viral lineage that jumped across the border from Gu\'{e}ck\'{e}dou prefecture in Guinea into Kailahun district in Sierra Leone.
The common ancestor of two of those (`14859\_EMLK' and `MK3462') continued onto Kenema district in Sierra Leone from Kailahun district (also Sierra Leone).
Unlike the lineage/transmission chain that eventually gave rise to strain `14859\_EMLK' which continued onto Conakry prefecture in Guinea and back-spilled into Sierra Leone's Kambia district (where the descendent strain `14859\_EMLK' was collected \cite{arias_rapid_2016}) right across the border later in the epidemic, the lineage/transmission chain that was ancestral to `MK3462' stayed in Sierra Leone for the remainder of the epidemic and found itself moved westwards towards Freetown (Western Urban and Western Rural districts) until finally jumping to Bombali district where its descendent strain was collected from a patient \cite{arias_rapid_2016}.
Unlike all the other three lineages the transmission chain/lineage (``lineage A'' \cite{carroll_temporal_2015}) that would leave strain `PL5294' as a descendent is thought to have been largely restricted to the environs of Conakry prefecture in Guinea.
An older ancestor of strain `PL5294' had migrated from Gu\'{e}ck\'{e}dou prefecture into Conakry prefecture on the other side of the country relatively early in the epidemic where its descendents circulated for a large portion of the West African epidemic though unlike a lot of its relatives the direct ancestor that would eventually give rise to strain `PL5294' spilled over into Kambia district of Sierra Leone \cite{arias_rapid_2016}.
Finally, the transmission chain/lineage that would eventually give rise to strain `EM\_004422' had a tumultous history in the region.
Strains `EM\_004422', `14859\_EMLK' and `MK3462' all shared the same common ancestor in Kailahun district of Sierra Leone but the progenitor of strain `EM\_004422', unlike its relatives, made a jump to Liberia (Lofa and Montserrado counties) from where \textit{its} descendants spilled back into Macenta prefecture in Guinea.
This transmission chain would later leave descendants (of which `EM\_004422' is representative) that jumped to neighbouring Kissidougou prefecture (also Guinea) and where strain `EM\_004422' was collected \cite{carroll_temporal_2015}.

The histories of the lineages that gave rise to these four tips are for the most part reconstructed from both GP sequences and genomes consistently (Figure \ref{trace}) likely as a result of additional information brought in by specifying tip dates and their collection locations.
Genomic data tend to concentrate the probability mass towards a single location at any given time in contrast to GP sequences where several locations can be considered with non-negligible probabilities at numerous time points (Figure \ref{trace} and Figure \ref{trace_entropy}) and where timing of ancestral migration events is considerably more diffuse or even substantially off (\textit{i.e.} Figure \ref{trace}A and D).
What is even more apparent is that without the additional information available when using complete genomes and without aiding the sampling with strongly informative priors MCMC explores a wider variety of low-probability migration paths as indicated by maps on the right of each plot in Figure \ref{trace}.
In the case of the ancestral lineage of `EM\_004422', for example, a series of migrations through distant Conakry (western Guinea) are reconstructed with relatively high confidence from GP sequences compared to shorter distance migrations that run through neighbouring Liberia reconstructed from genomes.
